---
title: "Introduction to visual.kaito"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to visual.kaito}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_deps <- requireNamespace("plotly", quietly = TRUE) &&
  requireNamespace("htmlwidgets", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = has_deps
)
```

## Statement of need

Comparing groups on more than one continuous outcome at a time is routine in
psychology and other social and health sciences -- for example, comparing a
clinical and a control group on two related symptom scales, or comparing
several treatment arms on two outcome measures simultaneously. The standard
R workflow for this either tests each variable separately (losing the joint,
multivariate picture) or reports a multivariate test (Hotelling's
T-squared, MANOVA) as a single number, with no visualization of *what the
joint distributions actually look like* or *how much they overlap*.

**visual.kaito** addresses this gap by drawing the two (or three) variables'
joint distributions directly as interactive 3D surfaces, with the visual
overlap between groups tied to the same statistics being reported (t-tests,
Hotelling's T-squared, one-way MANOVA, and four post-hoc procedures), rather
than as a separate, disconnected chart.

## Comparison with existing tools

Several excellent R packages already combine plots with statistical results
for group comparisons, but none currently draw the interactive 3D joint
distributions that **visual.kaito** focuses on:

* **ggstatsplot** and **ggpubr** overlay t-test/ANOVA/Kruskal-Wallis results
  and post-hoc annotations on standard 2D `ggplot2` charts (one variable at
  a time).
* **rstatix** provides a tidy interface to the underlying statistical
  tests (t-test, ANOVA, post-hoc comparisons) but no plotting of its own.
* **effectsize** computes and reports effect sizes, again without a
  plotting layer.
* **superb** focuses on summary/error-bar plots for factorial designs.

**visual.kaito** is not a replacement for these tools -- for a single
variable, a violin or box plot from `ggstatsplot` is usually the right
choice. It is meant for the specific case of *two or three continuous
variables compared jointly*, where seeing the overlap between groups'
distributions in 3D adds information that a table of per-variable p-values
does not.

## Installation

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("gygpsicologos-eng/visual.kaito")
```

```{r setup}
library(visual.kaito)
```

## Comparing two groups on two variables: `ttest_plot3d()`

`ttest_plot3d()` draws each group as a single bivariate density "mountain"
over two continuous variables, shades the region where the two mountains
overlap, and reports three complementary results: a Student's t-test for
each variable separately, and a joint Hotelling's T-squared test that
accounts for the correlation between the two variables.

```{r ttest-example}
set.seed(1)
df2 <- rbind(
  data.frame(iq = rnorm(30, 100, 12), anxiety = rnorm(30, 45, 8), group = "Control"),
  data.frame(iq = rnorm(30, 96, 11), anxiety = rnorm(30, 58, 9), group = "Clinical")
)

fig2 <- ttest_plot3d(df2, x = "iq", z = "anxiety", group = "group", n_grid = 35)
fig2
```

The full numeric results -- group means and covariances, both t-tests, and
the Hotelling's T-squared test -- are attached to the returned object and
do not require re-parsing the plot:

```{r ttest-stats}
str(attr(fig2, "stats"), max.level = 1)
```

The plot itself ships with four live controls (a view selector, an opacity
slider, a "what to report" selector, and a raw/standardized toggle) that
recompute instantly in the browser without calling R again -- useful for
exploring a result during a meeting or a class, not just for a static
figure.

## Comparing more than two groups: `manova_plot3d()`

When there are more than two groups, `manova_plot3d()` generalizes the same
idea: every group is drawn as its own bivariate density mountain, a
one-way MANOVA (Wilks' lambda) provides the omnibus test, and every pair of
groups can be compared post hoc using four different procedures --
Bonferroni, Tukey's HSD, Fisher's LSD ("DMS"), and Dunnett's test against a
user-chosen reference/control group.

```{r manova-example}
set.seed(2)
df3 <- rbind(
  data.frame(iq = rnorm(25, 100, 12), anxiety = rnorm(25, 45, 8), group = "Control"),
  data.frame(iq = rnorm(25, 96, 11), anxiety = rnorm(25, 58, 9), group = "Mild"),
  data.frame(iq = rnorm(25, 90, 13), anxiety = rnorm(25, 66, 10), group = "Severe")
)

fig3 <- manova_plot3d(df3, x = "iq", z = "anxiety", group = "group",
                       control = "Control", n_grid = 35)
fig3
```

Dunnett's test requires the optional 'multcomp' package. When it is not
installed, `manova_plot3d()` still draws the full plot with the other three
post-hoc methods; the Dunnett option shows an explanatory note instead of a
fabricated result.

## Triaxial box plots: `boxplot3d_interactive()`

For a more traditional (but still 3D and interactive) summary of three
continuous variables at once, `boxplot3d_interactive()` draws one
translucent 3D box per group, with whiskers, medians, raw points, and
outliers, using any of four whisker conventions (Tukey, fixed-percentile,
mean +/- SD, or letter-value):

```{r box-example}
set.seed(3)
df1 <- rbind(
  data.frame(iq = rnorm(30, 100, 12), anxiety = rnorm(30, 45, 8),
             mood = rnorm(30, 50, 10), group = "Control"),
  data.frame(iq = rnorm(30, 96, 11), anxiety = rnorm(30, 58, 9),
             mood = rnorm(30, 40, 11), group = "Clinical")
)

boxplot3d_interactive(df1, x = "iq", y = "anxiety", z = "mood", group = "group")
```

Live controls switch between the 3D scene and flat 2D panels for any pair
of axes, between whisker conventions, between raw and standardized scales,
and recolor the raw points by any grouping/categorical column in the data.
`boxplot3d_significance()` runs the matching statistical tests (per-axis
and joint, both parametric and non-parametric) without producing a plot,
for when only the numbers are needed.

## A note on interactivity

Every plot in this vignette is an interactive `htmlwidgets`/`plotly` object:
drag to rotate, scroll to zoom, and use the on-plot menus and sliders to
switch views. If you are reading a static (non-HTML) rendering of this
document, only a snapshot of the initial view will be visible -- build the
vignette locally (`browseVignettes("visual.kaito")`) or run the examples
directly to get the full interactive version.
